Using Metric Space Indexing for Complete and Efficient Record Linkage

### 2018-02-14 paper.tex
Before Final Revisions
After Final Revisions

\documentclass{llncs}
\usepackage{epsfig}
\usepackage{graphicx}

\newtheorem{mydef}{Def.}
\newtheorem{myhyp}{Hypothesis}

\usepackage{url}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{hhline}
\usepackage{caption}
\usepackage{subcaption}

%\pagestyle{empty}
\pagestyle{headings}

@@

\author{Submitted for double-blind review}

%  Hidden For Blind Review
\iffalse

\author{{\"O}zg{\"u}r Akg{\"u}n\inst{1} \and Alan Dearle\inst{1} \and\\
Graham Kirby\inst{1} \and Peter Christen\inst{2}}

\institute{School of Computer Science, University of St Andrews,\\
St Andrews, Scotland. Contact: \email{ozgur.akgun@st-andrews.ac.uk}
\and
Research School of Computer Science, The Australian National University,\\
Canberra, Australia. Contact: \email{peter.christen@anu.edu.au}}

\fi
%----------------------------------------------------------------------

@@

Record linkage is the process of identifying records that refer to the
same real-world entities, in situations where entity identifiers are
unavailable. Records are linked on the basis of similarity between
common attributes, with every pair being classified as a link or
non-link depending on their degree of similarity. Record linkage is
usually performed in a three-step process: first groups of similar
candidate records are identified using indexing, pairs within the same
group are then compared in more detail, and finally classified.
%
Even state-of-the-art indexing techniques, such as Locality Sensitive
Hashing, have potential drawbacks. They may fail to group together some
true matching records with high similarity. Conversely, they may group
records with low similarity, leading to high computational overhead.
%
We propose using metric space indexing to perform \emph{complete} record
linkage, which results in a parameter-free record linkage process
combining indexing, comparison and classification into a single step
delivering complete and efficient record linkage. Our experimental
evaluation on real-world datasets from several domains shows that
linkage using metric space indexing can yield better quality than
current indexing techniques, with similar execution cost, without the
need for domain knowledge or trial and error to configure the process.

@@

Record linkage, also known as entity resolution, data matching and
duplicate detection~\cite{Chr12}, is the process of identifying and
matching records that refer to the same real-world entities within or
across datasets. The entities to be linked are often people (such as
patients in hospital or customers in business datasets), but record
linkage can also be applied to link consumer products or bibliographic
records~\cite{Chr12}. Record linkage is commonly challenged by the lack
of unique entity identifiers (keys) in the datasets to be linked, which
prevents the use of a database join. Instead, the linkage of records
requires the comparison of the common attributes (or fields) that are
available within the datasets. For datasets that contain information
about individuals, these attributes include the names, addresses,
dates of birth, and so on, of individuals.

To overcome data quality issues such as typographical errors and
variations (which are common in name and address values~\cite{Chr12}),
approximate string comparison functions (such as edit distance, the
Jaro-Winkler comparator, or Jaccard similarity~\cite{Chr12}) are used to
compare pairs of records, leading to a vector of similarities (one
similarity per attribute compared) for each pair. These similarity
vectors are then used to classify the compared record pairs into
\emph{links} (where it is assumed both records in a pair correspond to
the same real-world entity) and \emph{non-links} (where the records are
assumed to correspond to two different entities). Various
classification methods have been employed in record
linkage~\cite{Chr12,Don15}, ranging from simple threshold-based to
sophisticated clustering, supervised classification techniques, and
active learning approaches~\cite{Wan15}.

Besides the issues of a lack of unique entity identifiers and data
quality (which will affect linkage quality), record linkage is also
challenged by the increasing sizes of the datasets to be
linked~\cite{Don15}. To avoid full pair-wise comparison of all possible
record pairs (quadratic in the sizes of the datasets to be linked),
blocking techniques, commonly known as \emph{indexing}~\cite{Chr12b},
are used. These split the datasets into smaller blocks in a
computationally efficient way, such that records that are likely to
correspond to the same entity are grouped into the same block. Only
records within the same block are then compared in more detail.

While indexing techniques facilitate efficient linkage of very large
datasets~\cite{Don15}, they generally achieve scalability at the cost
of reduced linkage quality, because potentially true matching record
pairs are removed in the indexing step, leading to a reduction in
recall~\cite{Chr12}. A variety of indexing
techniques, discussed in more detail in the following section, have
been proposed, ranging from simple phonetic based blocking~\cite{Chr12}
and sorting of the datasets~\cite{Dra12} to locality sensitive hashing
based techniques~\cite{Kim10,Steorts2014}. Techniques for
unsupervised~\cite{Kej13,Ram15} and supervised~\cite{Bil06,Mic06}
learning of optimal blocking schemes have also been proposed.

@@

Traditional record linkage systems that perform indexing prior to
comparison and classification, as illustrated on the left in
\Cref{fig-rl-process}, add a further practical complexity to the
process. Indexing, comparison and classification are often conducted
using algorithms and parameters selected based on technical and
domain expertise of the user of the record linkage system, followed
by a manual assessment (auditing) of the linkage
outcomes~\cite{Chr12}. If the quality of the resulting links is not
good enough for a certain application, the linkage process needs to
be repeated with different parameter settings and possibly also
alternative algorithms. This can result in a time-consuming and
labour-intensive iterative process. A major challenge is often the
heuristic nature of indexing, where the choice of a certain indexing
technique and its parameters (including which attributes to use in indexing) will determine the final outcome of a linkage.

This paper focuses on approaches that use a similarity threshold to
classify links. Such techniques are fundamentally limited by the extent
to which true matching records are similar, and true non-matches are
dissimilar---this is dataset-dependent. Within this domain, we define
a technique to be \emph{complete} if it is guaranteed to identify all
record pairs within the specified similarity threshold. Commonly used
indexing techniques are incomplete, since they reduce computational
cost at the expense of potentially overlooking some true matches. By
definition, incomplete techniques yield lower recall than complete
techniques. Conversely, and perhaps counter-intuitively, complete
techniques can yield lower precision with some datasets. This is
discussed further in Sect.~\ref{sec-approach}.

\emph{Metric space indexing} (MSI) is a complete technique with much
lower computational overheads than a brute force approach. It allows
indexing, comparison and classification to be combined into a single
step, as illustrated in the right-hand side in
Fig.~\ref{fig-rl-process}, making the overall record linkage process
simpler, more efficient and more effective than incomplete indexing
approaches.

\smallskip
\textbf{Contribution:} The primary contribution of this paper is the
novel application of MSI to achieve complete and efficient record
linkage, without the need for complex parameter tuning. We evaluate our
approach on several real-world datasets from diverse domains and
demonstrate its advantages over existing indexing techniques for
record linkage.

% For final paper:
% The motivation for this work is the Digitising Scotland project
% (Dibben 2012), which is in the process of transcribing and linking all
% the vital events recorded in Scotland between 1856 and 1977. This data
% set will, when complete, include around 14 million birth records, 11
% million death records and 4 million marriage records. As part of the
% work, certain data fields (locations, occupations and causes of death)
% are also being classified to the relevant standard coding schemes.

@@

We now briefly review relevant work in the areas of indexing for record
linkage (for recent surveys see~\cite{Chr12b,Pap16}), and metric space
indexing~\cite{Zezula2010}.

Techniques to link records across datasets have been investigated for
over five decades~\cite{Fel69,New59}, with the scalability of linking
being an ongoing challenge as datasets grow in size and complexity.
Traditional blocking~\cite{Chr12b} uses a set of attributes (known as a
\emph{blocking key}) to insert records that share the same value(s) in
their blocking key into the same block. Only records within the same
block are  compared to each other. To overcome variations and
misspellings, the values used in blocking keys can be phonetically
encoded using functions such as Soundex, NYSIIS, or
Double-Metaphone~\cite{Chr12}. These convert a string into a code
according to its pronunciation, assigning the same code to similar
sounding names (such as `Gail' and `Gayle'). Multiple blocking keys
may also be used to deal with the problem of missing attribute values.

A different approach to indexing is the sorted neighbourhood
method~\cite{Mon96}, where the datasets to be linked are sorted
according to a \emph{sorting key} (usually a concatenation of the values
from several attributes). A sliding window is then moved over the
datasets and only records within the window are compared. Techniques
that adaptively shrink or expand the window size based on the
characteristics of the sorting key values have been shown to improve
both linkage efficiency and quality~\cite{Dra12} over
%other sliding window
approaches that use fixed size windows.

These blocking techniques are heuristics, commonly requiring domain
knowledge, such as the choice of appropriate blocking or sorting keys.
Poor choices of blocking attributes result in records being inserted
into inappropriate blocks, and thus true matches being missed. As a
result, such techniques can lead to \emph{incomplete} linkage.
Conversely, many of the record pairs compared in a block may turn out
to have low similarity, corresponding to non-matches, resulting in
\emph{inefficient} linkage.

Locality sensitive hashing (LSH), originally proposed to allow efficient
nearest-neighbour search in high-dimensional spaces~\cite{Ind98}, has
been employed in record linkage as an indexing technique where attribute
values are hashed multiple times, and blocks are created from those
records that share some hash values. \emph{HARRA}~\cite{Kim10} is a
record linkage approach based on MinHash~\cite{Broder1997} and LSH
which blocks, compares, and then merges linked records in an iterative
fashion, where merged records are re-hashed to improve overall linkage
quality. \cite{Steorts2014} evaluates two LSH variations, concluding
that in order to get good results, LSH methods must be tuned to the
particular datasets being linked. This requires good quality ground
truth data which may be unavailable or expensive to obtain.

Metric space indexing (MSI) techniques~\cite{Zezula2010} provide
indexing structures to support the comparison of records in one set
with those in another, typically also offering similarity search
operations. To create an MSI data structure, it is necessary to define
a distance measure between records, with certain properties including
the \emph{triangle inequality}~\cite{Zezula2010}. Similarity search
operations include \emph{range-search(\textbf{q},d)}, where all
records $\mathbf{r}$ in a database $\mathbf{D}$ within a distance $d$
of a query record $\mathbf{q}$ are identified;
\emph{nearest-neighbour(\textbf{q})}, returning the record $\mathbf{r}
\in \mathbf{D}$ with smallest distance to $\mathbf{q}$; and
\emph{nearest-n(\textbf{q},n)}, returning the $n$ closest records in
$\mathbf{D}$ to $\mathbf{q}$.
%\cite{Zezula2010} gives an overview of MSI techniques.
Here we choose one metric space indexing structure, the
M-tree~\cite{Ciaccia97indexingmetric}, and investigate its efficacy
for record linkage.

The M-tree is a dynamically balanced tree structure. Every node contains
a reference to a record being indexed, a pointer to its parent, the
distance to its parent, and the node's radius. The radius of a node is
the distance from it to its furthest child. For a parent node with
radius \textit{r}, all its children may be visualised as being contained
within a ball of radius \textit{r} from it.

A threshold-based linkage method using R-trees~\cite{Hjaltason1998}
was described in~\cite{Li2006}. It demonstrates that high linkage
quality can be achieved using Jaccard similarity over selected record
attributes. \cite{Ciaccia97indexingmetric} shows that M-trees are
almost always more efficient than R-trees, hence their use in the
experiments described in this paper.

@@

We address the following general record linkage problem: for two
datasets $\mathbf{D}_A$ and $\mathbf{D}_B$, we wish to find, for each
record in $\mathbf{D}_A$, all the records in $\mathbf{D}_B$ that match
it with regard to a certain distance threshold, $d$ (i.e.\ have a
distance of $d$ or less). We compare several algorithms for record
linkage: traditional blocking, one incomplete similarity search method,
LSH-MinHash, and one complete method, M-tree. We also use a
simple complete brute force technique as a baseline for comparison,
although this can only feasibly be applied to the smallest of our
datasets. All experiments have a number of parameters to configure
the search space and algorithm behaviour, including a
\emph{distance threshold}, $d$, specifying a maximum distance,
equivalent to a minimum similarity, for two records to be classified
as a link (i.e.\ referring to the same entity).

\smallskip
\textbf{Brute force:}
Two nested loops are used to compare every record in
$\mathbf{D}_A$ with every record in $\mathbf{D}_B$. Each pair is
classified as a link if the distance between the records is equal to
or less than the threshold $d$. This approach is guaranteed to identify
all links, with complexity $ O(|\mathbf{D}_A| \times |\mathbf{D}_B|)$.

\smallskip
\textbf{Traditional blocking:}
The parameters of traditional blocking are the set of blocking keys
and the (optionally) phonetic
% Peter: normalisation?
encodings applied to each attribute. These are selected as described
in~\cite{Chr12b}, exploiting knowledge of the domain and of the
records being linked, and chosen with the intention of giving the
best possible results for the datasets. Each record in $\mathbf{D}_A$
is placed into the appropriate block based on its blocking key value.
The algorithm then iterates over the records in $\mathbf{D}_B$, and
for each one, compares it with each of the records from $\mathbf{D}_A$
in the block with the same blocking key value.

\smallskip
\textbf{LSH-MinHash:}
The parameters for LSH-Minhash are~\cite{Broder1997} \emph{shingle
size} ($l_{ss}$), \emph{band size} ($l_{bs}$) and \emph{number of
bands} ($l_{nb}$). Each record in $\mathbf{D}_A$ is placed in an
\emph{LSH-MinHash} data structure. First, all the attributes of the
record are concatenated, and the result \emph{shingled} into a set of
n-grams with $n = l_{ss}$. Next, a set of deterministically generated
hash functions are applied to each n-gram in the set and the smallest
result (the MinHash) of each hash application is added to a signature
for the record. The number of hashes used, and thus the size of the
signature, is set to $l_{nb} \times l_{bs}$. Finally, the signature is
split into $l_{nb}$ bands and the values from each band are hashed
again to create a number of keys. The original record is added to a
map associated with each of the keys.

To perform linkage, the algorithm iterates over the records in
$\mathbf{D}_B$. Each record is hashed using the procedure described
above, to obtain a set of keys. For each of these keys, the key is
looked up in the LSH-MinHash data structure, and the records from
$\mathbf{D}_A$ associated with it added to the result set. Finally,
the record from $\mathbf{D}_B$ is compared in turn with each record
in the result set using the same distance function and threshold, $d$,
as with the other approaches, with the pair being classified as a link
or non-link based on their distance.

In some circumstances, incomplete approaches such as traditional
blocking and LSH-MinHash can yield higher precision than complete
techniques. This can occur when the datasets contain a significant
number of non-matches that nonetheless have high similarity. In this
situation, the fact that an incomplete technique omits consideration
of some potential links can serve to improve precision, since a
classification decision based on a certain similarity threshold is
incorrect for high-similarity non-matches. By definition, recall can
never be higher for incomplete techniques.

\smallskip
\textbf{M-tree:} Besides the distance threshold, $d$, the M-tree
linkage algorithm has no additional parameters. In a similar manner
to the \emph{LSH-MinHash} approach, each record in $\mathbf{D}_A$ is inserted into an M-tree. To perform linkage, the algorithm iterates
over each record $\mathbf{b} \in \mathbf{D}_B$. A
\emph{range-search(\textbf{b},d)} operation is performed on the
M-tree, passing the distance threshold $d$ as the second parameter.
All the returned records are directly classified as links.

@@

We now describe the datasets we use in our evaluation, the set-up
employed to evaluate our proposed metric indexing approach and compare
it with traditional blocking techniques, and we then present and discuss
our results~\footnote{The raw data, additional figures and the source
code required to run these experiments can be downloaded from:
(link removed for double blind review).}.

@@

\begin{table}[t]
\caption{Characteristics of datasets used in the experiments.}
 \label{table-datasets}
  \centering
  \begin{scriptsize}
  %\addtolength{\tabcolsep}{-0.5pt}
  \begin{tabular}{ccccc}
  \hline\noalign{\smallskip}
  Dataset~ & ~Records in~& ~Records in~ & ~Number of true~& ~Entities \\
    name(s)  & ~dataset $\mathbf{D}_A$~ & ~dataset $\mathbf{D}_B$~ &
    matching pairs & linked \\
  \noalign{\smallskip} \hline \noalign{\smallskip}
  Cora & 1,295 & 1,295 & 17,184 & Publication--Publication\\
  ~Isle of Skye~ & 17,612 & 12,284& 2,900 & Birth--Death \\
  Kilmarnock & ~38,430~ & ~23,714~ & ~8,300~ & ~Birth--Death~ \\
  \noalign{\smallskip} \hline
  \end{tabular}
  \end{scriptsize}
\end{table}

@@

%\smallskip
%\textbf{Datasets:}
\label{sec-data}
We used three datasets from two domains in our experiments, as
summarised in Table~\ref{table-datasets}. The first is
\emph{Cora}~\cite{Cora2017}, which contains 1,295 records that refer
to 112 machine learning publications. Cora is commonly used as a
benchmark dataset in the literature for assessing linkage algorithms.
Ground truth is provided via a unique \emph{paper\_id} identifier of
the form ``blum1993". In this experiment linkage is performed over
the same set of records (i.e.\ a de-duplication~\cite{Chr12}).

The other two datasets are historical Scottish records of vital events
(birth, marriages and deaths) one registered on the \emph{Isle of
Skye}, a rural district, and the other records from
\emph{Kilmarnock}, an industrial town. These datasets were created by
historical demographers who extensively curated and linked both
datasets~\cite{reid2006,reid2002}. Both datasets include the names,
gender, addresses of individuals and their parents. Ground truth was
generated by the demographers based on their extensive domain
knowledge.

In all of our experiments we use a single distance metric which is the
sum of the Levenshtein~\cite{Levenshtein66} edit distances between the
set of attributes being compared.

@@

In this section, we perform linkage on the Cora dataset using all four
approaches presented in this paper: brute force, traditional blocking,
LSH and M-tree, using several selected configurations for the blocking
and LSH approaches. The distance threshold is varied between $0 \le d
\le 250$~\footnote{Relatively high Levenshtein edit distances are
included since Cora includes a number of low-similarity true
matches.}. For traditional blocking, the following Cora attributes are
used individually as blocking keys: \emph{author}, \emph{title},
\emph{venue}, \emph{location}, \emph{publisher} and \emph{year}. We
also use a combined blocking key comprising all attributes.

\Cref{cora-quality} shows the precision, recall, and F-measure~\cite{Chr12} for various thresholds. As expected, low 
thresholds give high precision and low recall, and the reverse for high
threshold values. Brute force and M-tree give identical results, as
expected. The best linkage quality, with an F-measure of around 0.7,
is achieved by a number of linkers, including brute force, M-tree, blocking on \emph{authors} and blocking on all attributes, and two
of the LSH configurations. All of these give similar overall results,
apart from blocking on \emph{authors}, which gives much better quality
at very high distance thresholds. This is due to the incomplete
nature of the approach, avoiding comparisons of significant numbers
of high-similarity non-matches and thus avoiding these becoming false
positives and keeping precision high.

@@

The observation that LSH can yield results quickly at the cost of some
recall suggests that hybrid algorithms may be a potentially profitable
future direction for exploration. We hypothesise that it may be possible
to use an incomplete technique like LSH to form indices into a data
structure such as an M-tree. In this manner, M-tree nodes could be used
as the starting point for range-search queries and, in so doing, prune
the search space considerably to yield high quality results quickly.
We plan to develop such hybrid indexing approaches with the aim to
achieve fully automated, complete and efficient record linkage.

>>>>

\documentclass{llncs}

\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{url}
\usepackage{cleveref}
\usepackage{hhline}
\usepackage{subcaption}

@@

\author{{\"O}zg{\"u}r Akg{\"u}n\inst{1} \and Alan Dearle\inst{1}
  \and Graham Kirby\inst{1} \and Peter Christen\inst{2}}

\institute{School of Computer Science, University of St Andrews,
St Andrews, Scotland. Contact: \email{\{ozgur.akgun, alan.dearle, graham.kirby\}@st-andrews.ac.uk}
\and
Research School of Computer Science, The Australian National University,\\
Canberra, Australia. Contact: \email{peter.christen@anu.edu.au}}

@@

Record linkage is the process of identifying records that refer to the
same real-world entities, in situations where entity identifiers are
unavailable. Records are linked on the basis of similarity between
common attributes, with every pair being classified as a link or
non-link depending on their similarity. Linkage is
usually performed in a three-step process: first, groups of similar
candidate records are identified using indexing, then pairs within the same
group are compared in more detail, and finally classified.
Even state-of-the-art indexing techniques, such as Locality Sensitive
Hashing, have potential drawbacks. They may fail to group together some
true matching records with high similarity, or they may group
records with low similarity, leading to high computational overhead.
We propose using \emph{metric space indexing} (MSI) to perform \emph{complete}
linkage, resulting in a parameter-free process
combining indexing, comparison and classification into a single step
delivering complete and efficient record linkage. An
evaluation on real-world data from several domains shows that
linkage using MSI can yield better quality than
current indexing techniques, with similar execution cost, without the
need for domain knowledge or trial and error to configure the process.

@@

Record linkage, also known as entity resolution, data matching and
duplicate detection~\cite{Chr12}, is the process of identifying and
matching records that refer to the same real-world entities within or
across datasets. The entities to be linked are often people (such as
patients in hospital or customers in business datasets), but record
linkage can also be applied to link consumer products or bibliographic
records~\cite{Chr12}. Record linkage is commonly challenged by the lack
of unique entity identifiers (keys) in the datasets to be linked, which
prevents the use of a database join. Instead, the linkage of records
requires the comparison of the common attributes (or fields) that are
available within the datasets, for example the names, addresses and
dates of birth of individuals.

To overcome data quality issues such as typographical errors and
variations (common in name and address values~\cite{Chr12}),
approximate string comparison functions (e.g. edit distance, the
Jaro-Winkler comparator, or Jaccard similarity~\cite{Chr12}) are used to
compare record pairs, leading to a vector of similarities (one
similarity per attribute compared) for each pair. These
are used to classify the record pairs into
\emph{links} (where it is assumed both records correspond to
the same real-world entity) and \emph{non-links} (where they are
assumed to correspond to different entities). Various
classification methods have been employed in record
linkage~\cite{Chr12,Don15}, ranging from simple threshold-based to
sophisticated clustering, supervised classification techniques, and
active learning approaches~\cite{Wan15}.

Besides a lack of unique entity identifiers, and data
quality issues, linkage is also
challenged by dataset scale~\cite{Don15}. To avoid full pair-wise comparison of all possible
record pairs (quadratic in the dataset sizes),
blocking techniques, commonly known as \emph{indexing}~\cite{Chr12b},
are used. These split the datasets into smaller blocks in an efficient way, grouping together records that are likely to
correspond to the same entity. Only
records within blocks are then compared in detail.

While indexing allows efficient linkage of large
datasets~\cite{Don15}, scalability is at the cost of
reduced linkage quality, because potentially matching record pairs
are ignored, leading to lower
recall~\cite{Chr12}. Indexing techniques, discussed in more
detail later, range from simple
phonetic based blocking~\cite{Chr12} and sorting of the
datasets~\cite{Dra12} to locality sensitive hashing based
techniques~\cite{Kim10,Steorts2014}, and unsupervised~\cite{Kej13,Ram15} and supervised~\cite{Bil06,Mic06}
learning of optimal blocking schemes.

@@

Traditional linkage systems that perform indexing prior to
comparison and classification (on the left in
\Cref{fig-rl-process}) add a further complexity. Indexing, comparison and classification are often conducted
using algorithms and parameters selected using
domain expertise, followed
by manual assessment of the linkage
outcomes~\cite{Chr12}. If the resulting link quality is too low for a certain application, the process is
repeated with different parameter settings or algorithms, giving a time-consuming iterative process. The choice of an appropriate indexing
technique and parameters (including which attributes to use in
indexing) will significantly affect the final linkage outcome.

We focus on approaches using a similarity threshold to
classify links. These are fundamentally limited by the extent
to which true matching records are similar, and true non-matches are
dissimilar---this is dataset-dependent. Within this domain, we define
a technique to be \emph{complete} if it guarantees to find all
record pairs within the specified threshold. Many
indexing techniques are incomplete, since they reduce computational
cost at the expense of potentially overlooking some true matches. By
definition, incomplete techniques yield lower recall than complete
ones. Conversely, and counter-intuitively, complete
techniques can yield lower precision with some datasets. This is
discussed further in Sect.~\ref{sec-approach}.

Metric space indexing (MSI) is a complete technique with
lower computational cost than a brute force approach. It allows
indexing, comparison and classification to be combined into a single
step (on the right in
\Cref{fig-rl-process}), making the process
simpler, more efficient and more effective than incomplete
approaches.

\textbf{Contribution:} Our primary contribution is the
novel application of MSI to achieve complete and efficient record
linkage, without the need for complex parameter tuning. We evaluate our
approach on several real-world datasets and
demonstrate its advantages over existing indexing techniques for
record linkage~\footnote{The motivation for this work is the Digitising Scotland
project~\cite{Dibben2012}, which aims to transcribe and
link all civil registration events recorded in Scotland between
1856 and 1973. This data set will include around 14
million birth records, 11 million death records and 4 million marriage
records.}.

@@

We review relevant work in the areas of indexing for record
linkage (for recent surveys see~\cite{Chr12b,Pap16}), and metric space
indexing~\cite{Zezula2010}. Techniques to link records have been investigated for
over five decades~\cite{Fel69,New59}, with scalability
being an ongoing challenge as datasets grow in size and complexity.
Traditional blocking~\cite{Chr12b} uses a set of attributes (a
\emph{blocking key}) to insert records with the same value(s) in
their blocking key into the same block. Only records within the same
block are  compared to each other. To overcome variations and
misspellings, the values can be phonetically
encoded using functions such as Soundex, NYSIIS, or
Double-Metaphone~\cite{Chr12}. These convert a string into a code
according to its pronunciation, assigning the same code to similar
sounding names (such as `Gail' and `Gayle'). Multiple blocking keys
may also be used to deal with missing attribute values.

A different approach uses sorted neighbourhoods~\cite{Mon96}, where the datasets are sorted
according to a \emph{sorting key} (usually a concatenation of several attribute values). A sliding window is moved over the
datasets and only records within the window are compared. Techniques
that adaptively shrink or expand the window size based on the
characteristics of the sorting key values have been shown to improve
both linkage efficiency and quality~\cite{Dra12}.

These techniques are heuristics, requiring domain
knowledge, such as the choice of appropriate blocking or sorting keys.
Poor choices of blocking attributes result in records being inserted
into inappropriate blocks, and thus true matches being missed, giving \emph{incomplete} linkage.
Conversely, many pairs compared in a block may have low similarity, being non-matches, giving
\emph{inefficient} linkage.

Locality sensitive hashing (LSH), proposed for efficient
nearest-neighbour search in high-dimensional spaces~\cite{Ind98}, has
been used for record linkage indexing. Attribute
values are hashed multiple times, and blocks are created from those
records that share some hash values. \emph{HARRA}~\cite{Kim10} is a
linkage approach based on MinHash~\cite{Broder1997} and LSH
which blocks, compares, and then merges linked records iteratively. \cite{Steorts2014} evaluates two LSH variations, concluding
that to get good results, they must be tuned to the
datasets. This requires good ground
truth data which may be unavailable or expensive to obtain.

Metric space indexing (MSI) techniques~\cite{Zezula2010} support similarity search. 
They require
a distance measure between records, with certain properties including
the \emph{triangle inequality}~\cite{Zezula2010}. Similarity search
operations include \emph{range-search(\textbf{q},d)}, identifying all
records within a distance $d$
of a query record $\mathbf{q}$;
\emph{nearest-neighbour(\textbf{q})}, returning the record with smallest distance to $\mathbf{q}$; and
\emph{nearest-n(\textbf{q},n)}, returning the $n$ closest records to $\mathbf{q}$. Here we choose one MSI
structure, the M-tree~\cite{Ciaccia97indexingmetric}, and investigate
its efficacy for record linkage.
The M-tree is dynamically balanced. Every node contains
a reference to a record being indexed, a pointer to its parent, the
distance to its parent, and the node's radius. The radius of a node is
the distance from it to its furthest child. For a parent node with
radius \textit{r}, all its children may be visualised as being contained
within a ball of radius \textit{r} from it.

A linkage method using R-trees~\cite{Hjaltason1998}
was described in~\cite{Li2006}, demonstrating that high linkage
quality can be achieved using Jaccard similarity. \cite{Ciaccia97indexingmetric} shows that M-trees are
almost always more efficient than R-trees, hence their use here.

@@

We address the following general linkage problem: for two
datasets $\mathbf{D}_A$ and $\mathbf{D}_B$, we wish to find, for each
record in $\mathbf{D}_A$, all the records in $\mathbf{D}_B$ that match
it with regard to a certain distance threshold $d$ (i.e.\ have a
distance of $d$ or less). We compare several linkage algorithms: traditional blocking, an incomplete similarity search method,
LSH-MinHash, and a complete method, M-tree. We also use a
complete brute force technique as a baseline, though
this can only feasibly be applied to our smallest dataset. All
experiments have a number of parameters to configure the search space
and algorithm behaviour, including the distance function and the threshold, $d$,
specifying the maximum distance for
two records to be classified as a link (i.e.\ referring to the same
entity).
We focus on a single distance function in these experiments, to constrain the experimental space. In
Sect.~\ref{sec-concl} we return to the selection of alternative distance functions.

\textbf{Brute force:}
Every record in
$\mathbf{D}_A$ is compared with every record in $\mathbf{D}_B$. Each pair is
classified as a link if the distance between the records is less than or equal to
the threshold. This always finds
all links, with complexity $ O(|\mathbf{D}_A| \times |\mathbf{D}_B|)$.

\textbf{Traditional blocking:}
The parameters are the set of blocking keys and
(optionally) the phonetic encodings applied to each attribute. These are
selected as described in~\cite{Chr12b}, exploiting knowledge of the
domain and of the data, and chosen with the intention of
giving the best possible results. Each record in
$\mathbf{D}_A$ is placed into the appropriate block based on its
blocking key value. The algorithm then iterates over the records in
$\mathbf{D}_B$, and for each one, compares it with each of the records
from $\mathbf{D}_A$ in the block with the same blocking key value.

\textbf{LSH-MinHash:}
The parameters for LSH-Minhash are~\cite{Broder1997} \emph{shingle
size} ($l_{ss}$), \emph{band size} ($l_{bs}$) and \emph{number of
bands} ($l_{nb}$). First, the attributes of each record in $\mathbf{D}_A$
are concatenated, and the result \emph{shingled} into a set of
n-grams with $n = l_{ss}$. Next, a set of deterministically generated
hash functions are applied to each n-gram in the set and the smallest
result (the MinHash) of each hash application is added to a signature
for the record. The number of hashes used, and thus the size of the
signature, is set to $l_{nb} \times l_{bs}$. Finally, the signature is
split into $l_{nb}$ bands and the values from each band are hashed
again to create a number of keys. The original record is added to a
map associated with each of the keys. 
To perform linkage, the algorithm iterates over the records in
$\mathbf{D}_B$. Each record is hashed as described
above, to obtain a set of keys. Each key is
looked up in the data structure, and the associated records from
$\mathbf{D}_A$ added to the result set. Finally,
the record from $\mathbf{D}_B$ is compared in turn with each record
in the result set, with the pair being classified as a link
or non-link based on their distance.

In some circumstances, incomplete approaches such as traditional
blocking and LSH-MinHash can yield higher precision than complete
techniques. This can occur when a significant
number of non-matches nonetheless have high similarity. In this
situation, the fact that an incomplete technique omits consideration
of some potential links can serve to improve precision, since a
classification decision based on a certain similarity threshold is
incorrect for high-similarity non-matches. By definition, recall can
never be higher for incomplete techniques.

\textbf{M-tree:} The linkage
algorithm has no additional parameters. As with
\emph{LSH-MinHash}, each record in $\mathbf{D}_A$ is inserted
into an M-tree. To perform linkage, the algorithm iterates over each
record $\mathbf{b} \in \mathbf{D}_B$. A
\emph{range-search(\textbf{b},d)} operation is performed on the M-tree,
passing the distance threshold $d$ as the second parameter. All the
returned records are directly classified as links.

@@

We now describe the datasets and method used in our evaluation~\footnote{The raw data, additional figures and the source
code required to run these experiments can be downloaded from:
(link removed for double blind review).}.
\label{sec-data}
We used three datasets from two domains in our experiments, as
summarised in Table~\ref{table-datasets}. The first is
\emph{Cora}~\cite{Cora2017}, which contains 1,295 records that refer
to 112 machine learning publications. Cora is commonly used as a
benchmark dataset in the literature for assessing linkage algorithms.
Ground truth is provided via a unique \emph{paper\_id} identifier of
the form ``blum1993". In this experiment linkage is performed over
the same set of records (i.e.\ a de-duplication~\cite{Chr12}).

@@

\begin{table}[t]
\caption{Characteristics of datasets used in the experiments.}
 \label{table-datasets}
  \centering
  \begin{scriptsize}
  \begin{tabular}{ccccc}
  \hline\noalign{\smallskip}
  Dataset~ & ~Records in~& ~Records in~ & ~Number of true~& ~Entities \\
    name(s)  & ~dataset $\mathbf{D}_A$~ & ~dataset $\mathbf{D}_B$~ &
    matching pairs & linked \\
  \noalign{\smallskip} \hline \noalign{\smallskip}
  Cora & 1,295 & 1,295 & 17,184 & Publication--Publication\\
  ~Isle of Skye~ & 17,612 & 12,284& 2,900 & Birth--Death \\
  Kilmarnock & ~38,430~ & ~23,714~ & ~8,300~ & ~Birth--Death~ \\
  \noalign{\smallskip} \hline
  \end{tabular}
  \end{scriptsize}
\end{table}

@@

The other two datasets are historical Scottish records of vital events
(birth, marriages and deaths), one registered on the \emph{Isle of
Skye}, a rural district, and the other records from
\emph{Kilmarnock}, an industrial town. These datasets were created, curated and linked by
historical demographers~\cite{reid2006,reid2002}. Both include the names and
genders of individuals and their parents. Ground truth was
generated by the demographers based on their extensive domain
knowledge.

In all of our experiments we use a single distance metric: the
sum of the attribute-level Levenshtein~\cite{Levenshtein66} edit distances.

@@

We perform linkage on the Cora dataset using all
approaches presented in this paper: brute force, traditional blocking,
LSH and M-tree, using several selected configurations for blocking
and LSH. The distance threshold is varied between 0 and 250~\footnote{Relatively high Levenshtein edit distances are
included since Cora includes a number of low-similarity true
matches.}. For traditional blocking, the following attributes are
used individually as blocking keys: \emph{author}, \emph{title},
\emph{venue}, \emph{location}, \emph{publisher} and \emph{year}. We
also use a combined blocking key comprising all attributes.

\Cref{cora-quality} shows the precision, recall, and
F-measure~\cite{Chr12} for various thresholds. As expected, low
thresholds give high precision and low recall, and the reverse for high
thresholds. Brute force and M-tree give identical results, as
expected. The best linkage quality, with an F-measure of around 0.7, is
achieved by several linkers, including brute force, M-tree, blocking
on \emph{authors}, blocking on all attributes, and two of the LSH
configurations. All of these give similar overall results, apart from
blocking on \emph{authors}, which gives much better quality at very high
distance thresholds. This is due to the incomplete nature of the
approach, avoiding comparisons of significant numbers of high-similarity
non-matches and thus avoiding these becoming false positives and keeping
precision high.

@@

We have focused on a single distance function: the sum of attribute-level Levenshtein distances. This gives a straightforward intuition of record-level distances, but it is more common to normalise metrics to the range 0-1.

Many distance functions, such as Jaro-Winkler, are not metric, and therefore cannot be used with MSI techniques. Care must be taken to preserve metric properties when combining metrics over individual fields, as is highlighted in~\cite{Yujian2007}, since this may yield a function that is not metric. It is then possible for MSI techniques to yield results that are subtly incorrect.

Different metrics can give different distributions of inter-record distances, which can affect both the linkage results and the number of comparisons made, and hence the efficiency of the algorithm.
Datasets with low variation in inter-record distances are said to have high \textit{intrinsic dimensionality}, and tend to require high numbers of comparisons.
The Scottish vital event datasets, combined with Levenshtein-based metrics, appear to have high intrinsic dimensionality.
We are now, therefore, investigating the application of various different metrics to this domain, including Jensen-Shannon, Cosine and Structured Entropic Distance, as described in~\cite{Connor2016}. We are also investigating the applicability of a novel technique for dimensionality reduction~\cite{Connor2017}.

\section{Acknowledgements}
